Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-13 content.tex
Updates from Overleaf

  \textbf{1. }
    Phylogenetic trees are current routinely reconstructed from an alignment of character sequences (usually nucleotide sequences). Bayesian tools, such as MrBayes, RevBayes and BEAST2, have gained much popularity over the last decade, as they allow joint estimation of the posterior distribution of the phylogenetic tree and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree prior. While in principle the Bayesian framework allows for comparing different species tree priors and hence may elucidate the macroevolutionary processes underlying the species tree, in practice only 
    macroevolutionary models that allow for fast computation of the prior probability are used. 
    This begs the question how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in the tree prior. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of BEAST2, but its philosophy applies to any Bayesian phylogenetic inference tool\\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their interpretation. \\
\end{abstract}

@@

The development of new powerful Bayesian phylogenetic inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
has been a major advance in building phylogenetic trees 
from character data (usually nucleotide sequences) extracted 
from extant (but also extinct) organisms, and hence in our understanding of the main drivers and modes of diversification.

BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
that needs both character data and priors to infer a posterior.
Specifically, the species tree prior describes the process of diversification.
BEAST has built-in tree priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing 
functionalities.
Hence one can add novel diversification models by writing a BEAST2 package that contains the likelihood formula
of a phylogeny under the novel diversification model, i.e. the prior probability of a species tree.

Many such diversification models (and their associated probability 
algorithms) have been developed
developed, e.g., those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or where diversification rates change for specific lineages and their descendants [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}], models that treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}], or where diversification rate
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only few of these diversification models are available as a BEAST2 package.

When a novel diversification model is introduced,
its performance in inference is commonly tested.
Part of a model's performance is its ability to recover parameters from simulated data with known parameters (e.g. [\cite{etienne2014estimating}]),
Ideally, the estimated parameter values closely match the known/true values.

@@

Here we introduce a method to quantify the importance of a novel tree prior.
The method starts with a phylogeny generated by the new model for the tree prior. Next, nucleotide sequences are simulated on this phylogeny. Then, using the tree priors built-in into BEAST2, a Bayesian posterior distribution of phylogenies is inferred. We can then compare the inferred and simulated phylogenies. How to properly perform this comparison forms the heart of our method.
Only new diversification models that result in a large discrepancy between inferred and simulated phylogeny will be worth the effort and computational burden 
to implement a species tree prior for in a Bayesian framework.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior' $\mathit{p_{G}}$.
Many tree priors have a parameter setting in which these are equivalent
to a standard tree prior. For example, a protracted birth-death model reduces to a standard birth-death model when speciation-completion rate approaches infinity [i.e. rate $\lambda$ in \cite{etienne2014estimating}, eqs. (2b) and (2c)].
When benchmarking a novel tree prior, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations when assessing a novel tree prior, our examples contain only one replicate as they are for illustrative purposes only.

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in tables~\ref{tab:options} and~\ref{tab:definitions}.

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer. 
These ERQs gradually build up in complexity.
For each of these ERQs we will quantify the importance of a tree prior.

@@

The pipeline is summarized by the following steps:
\begin{enumerate}
  \item from a given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is picked (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a (DNA) root sequence, 
a (DNA) mutation rate, a clock model and a nucleotide substitution
model, which will be referred to as site model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is supported in this step.

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we can express that an inference
model (a combination of a tree prior, clock model and site model) 
is known to include the generative/true tree prior and that 
we definitely want to use that model in our inference. 
As a second example, we can also express that we've submitted a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, where in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when being the best.
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. 
Scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 4a),
and those inference models that were selected in the previous step (3). 
For each selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package which makes use of BEAST2. 
This step usually takes up most of the full pipeline's computation time.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
A phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Assume that a non-standard generative prior $\mathit{p_{G}}$
produces a tree $\tau_{\mathit{G}}$,
with branching times $\Vec{t}_{\mathit{G}}$ and 
topology $\psi_{\mathit{G}}$.
The twinning process T, in accordance with a chosen inference 
model $\mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving the 
topology $\psi_{\mathit{G}}$:

@@

The left-over differences in error distributions will be 
consequences of different phylogenies and tree priors for the two parallel 
pipelines: the main pipeline uses the tree $\tau_{\mathit{G}}$ and a
possibly mismatched prior $\mathit{G}$, where the twin pipeline
uses the tree $\tau_{\mathit{I}}$ for a matching prior $\mathit{I}$.
So, for the twin route, if the process is repeated a sufficient number of times, the resulting error distribution cannot be 
attributed to a mismatch between the true tree prior and the tree prior used in
inference.

@@

\verb;pirouette; will be made available on CRAN from which it can then be easily installed easily from CRAN:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single
]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}

To start using \verb;pirouette;, load its functions in the global namespace 
first:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select the best candidate model,
\verb;pirouette; needs the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

An overview of \verb;pirouette;'s main functions is shown in 
table~\ref{tab:functions}. 
Their usage is demonstrated in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example and sensible defaults.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists in four blocks of 250 
mononucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all species are equal and constant.

@@

As the site and clock models used here are also the defaults, 
these function arguments can be safely omitted: we just show these
explicitly for clarity.

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made from a phylogeny 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

The resulting error distribution, as shown figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, 
thus cannot be part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the user has to somehow select 
a standard tree prior to be 
used in the inference.
There are two ways to deal with this problem: hand-picking a tree
prior, or let the best of a set of models be picked. 
Hand-picking a tree prior can be rational if one standard tree prior 
is more suitable than others for the inference, for example, when
the novel tree prior is most similar to a certain tree prior.
Instead of hand-picking, the best inference model from a list of candidate 
models can be used in inference.
\verb;pirouette; selects the 'best' model, that is, 
the model with the highest evidence (also known as marginal likelihood).

@@

In the second step, unlike ERQ 1, we provide a 
set of experiments $\{\mathit{X_{i}}\}_{i=1}^{40}$. All the experiments have 
the same inference conditions $\mathit{C_{i}} = \mathit{C}$ but different 
candidate inference models $\mathit{I_{i}}$. Only one of them will be used 
to run the actual inference. According to the specified $\mathit{C}$, 
\verb;pirouette; will measure the evidence for each of them and select only 
the one with the highest evidence as the best candidate.

As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function to provide it:
\verb;create_all_experiments; will create a full set of 40 experiments,
containing the inference models of all combinations of 
4 site models, 2 clock models and 5 tree priors.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
as shown in figure~\ref{fig:example_2}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    ERQ 2: the inference error from
    when selecting the best inference model.
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}
\giovanni{We want to import the "twin\_model" from the inference model.}
\richel{Yes, that is an option, but according to Rampal not important enough
to add now}

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
The output is shown in figure~\ref{fig:example_3}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    ERQ 3: the inference error made 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

@@

When no inference model can be hand-picked, 
one can select the best model from a set,
as shown in ERQ2.
The resulting error distribution contains the error accumulated in the pipeline,
without quantifying to what extent this error is due to the mismatch between 
the generative model $\mathit{G}$ and the selected inference model 
$\mathit{I}$.
As twinning (see section \ref{subsec:twinning}: 'Twinning') shows us the baseline error, we can quantify the contribution of the tree prior to the error.

@@

We specify that we want all 40 inference models to compete 
by calling \verb;create_all_experiments;,
as we previously did in listing~\ref{lst:create_all_experiments}.
We choose again to use the default error measurement setup,
as shown in listing~\ref{lst:create_error_measure_params}.
To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using
the same call to \verb;create_pir_params; as shown in 
listing~\ref{lst:create_pir_params_with_twinning}.
We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in figure~\ref{fig:example_4}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative 
    model is known, compared to candidates with similar inference models, in 
    relationship to the background error.
  }
  \label{fig:example_4}
\end{figure}

@@

We choose again to use the default error measurement parameters
created in listing~\ref{lst:create_error_measure_params}.
We enable twinning, by calling \verb;create_pir_params;
as shown in listing~\ref{lst:create_pir_params_with_twinning}.

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in figure~\ref{fig:example_5}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the inference error made, 
    when the generative model is hand-picked, 
    compared to the best of 39 candidate models, 
    each in relationship to the background error.
  }
  \label{fig:example_5}
\end{figure}

@@

In the five example research questions, 
we showed how to use \verb;pirouette; 
to quantify the importance of a tree prior in Bayesian phylogenetics.
In all ERQs we used and assumed the most simple tree prior possible,
which is the Yule tree prior.
In principle any other (more complex) generative tree prior 
can be tested, but we chose to provide the 
simplest (and fastest to run) examples.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.

>>>>

  \textbf{1. }
    Phylogenetic trees are current routinely reconstructed from an alignment of character sequences (usually nucleotide sequences). Bayesian tools, such as MrBayes, RevBayes and BEAST2, have gained much popularity over the last decade, as they allow joint estimation of the posterior distribution of the phylogenetic tree and the parameters of the underlying inference model.  
    An important ingredient of these Bayesian approaches is the species tree prior. While in principle the Bayesian framework allows for comparing different species tree priors and hence may elucidate the macroevolutionary processes underlying the species tree, in practice only 
    macroevolutionary models that allow for fast computation of the prior probability are used. 
    This begs the question how accurate the tree estimation is
    when the real macroevolutionary processes are substantially different 
    from those assumed in the tree prior. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error made by Bayesian phylogenetics for a given macroevolutionary diversification model. \verb;pirouette; makes use of BEAST2, but its philosophy applies to any Bayesian phylogenetic inference tool\\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples and their interpretation. \\
\end{abstract}

@@

The development of new powerful Bayesian phylogenetic inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}], 
has been a major advance in constructing phylogenetic trees 
from character data (usually nucleotide sequences) extracted 
from extant (but also extinct) organisms, and hence in our understanding of the main drivers and modes of diversification.

BEAST [\cite{drummond2007beast}] is a typical Bayesian phylogenetics tool, 
that needs both character data and priors to infer a posterior distribution of phylogenies.
Specifically, for the species tree prior - which describes the process of diversification - 
BEAST has built-in priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] models.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing functionalities. For example, one can add novel diversification models by writing a BEAST2 package that contains the likelihood formula
of a phylogeny under the novel diversification model, i.e. the prior probability of a species tree.
Many such diversification models (and their associated probability algorithms) have been developed, e.g., models in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or where diversification rates change for specific lineages and their descendants [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}], models that treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}][\cite{lambert2015reconstructed}], 
or as a burst of simultaneous branching
events [\cite{laudanno2018mbd}], or where diversification rate
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
even concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only a few of these diversification models are available as a BEAST2 package.

When a novel diversification model is introduced,
its performance in inference should be tested.
Part of a model's performance is its ability to recover parameters from simulated data with known parameters (e.g. [\cite{etienne2014estimating}]), where ideally the estimated parameter values closely match the known/true values.

@@

Here we introduce a method to quantify the importance of a novel tree prior.
The method starts with a phylogeny generated by the new model for the tree prior. Next, nucleotide sequences are simulated on this phylogeny. Then, using the tree priors built-in into BEAST2, a Bayesian posterior distribution of phylogenies is inferred. We can then compare the inferred and simulated phylogenies. How to properly perform this comparison forms the heart of our method.
Only new diversification models that result in a large discrepancy between inferred and simulated phylogenies will be worth the effort and computational burden 
to implement a species tree prior for in a Bayesian framework.

@@

\verb;pirouette; is written in the R programming language (\cite{R}).
The goal of \verb;pirouette; is to quantify the importance of a tree prior.
It does so by measuring the inference error made for a given reconstructed phylogeny, 
simulated under a (usually novel) diversification model.
We refer to the true prior that generated the given tree
as the 'generative tree prior' $\mathit{p_{G}}$.
Many tree priors have a parameter setting for which it reduces to a standard tree prior. For example, a protracted birth-death model reduces to a standard birth-death model when speciation-completion rate approaches infinity [i.e. rate $\lambda$ in \cite{etienne2014estimating}, eqs. (2b) and (2c)].
When benchmarking a novel tree prior, 
one will typically construct phylogenies 
for different combinations of the diversification model's parameters, 
to assess under which scenarios the inference error cannot be neglected. 
While we recommend many replicate simulations when assessing a novel tree prior, our examples contain only one replicate as they are for illustrative purposes only.

\verb;pirouette; is very flexible and allows the user 
to specify a wide variety of custom settings. 
These settings can be grouped in macro-sections, 
according to how they operate in the pipeline. 
We summarize them in Tables~\ref{tab:options} and~\ref{tab:definitions}.

Although many possible tests can be performed, 
we present here what we think are the five most important 
example research questions (ERQ) that \verb;pirouette; can answer, from simple questions to complex ones.
For each of these ERQs we will quantify the importance of a tree prior.

@@

We assume the user has a phylogeny simulated with the new diversification model. The pipeline to assess the error BEAST2 makes in inferring this phylogeny then contains the following steps:
\begin{enumerate}
  \item from the given phylogeny an alignment is simulated 
    under a known alignment model $\mathit{A}$;
  \item from this alignment, according to the specified inference conditions $\mathit{C}$, 
    an inference model $\mathit{I}$ is picked (which may differ from the 
    generative model);
  \item the inference model and the alignment are used 
    to infer a posterior distribution of phylogenies;
  \item the phylogenies in the posterior are compared with the given phylogeny 
    to estimate the error made, according to the error measure $\mathit{E}$ specified 
    by the user;
\end{enumerate}
The pipeline is visualized in Fig.~\ref{fig:pipeline}. 
There is also the option to generate a 'twin tree', 
that goes through the same pipeline. 
The utility of this twin tree will be explained below.

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of a (DNA) root sequence, a (DNA) mutation rate, a clock model and a nucleotide substitution
model, which will be referred to as site model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported site models are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

The second step (Fig.~\ref{fig:pipeline}, 3)
selects one or more inference models $I$ from a set of inference 
models $I_{1},\dots,I_{n}$. 
We define an experiment $X_{i}$ as the combination of 
an inference model $I_{i}$ and the conditions $C_{i}$ 
to actually use it in the inference step.
For example, we may require that an inference
model (a combination of a tree prior, clock model and site model) 
should include the generative/true tree prior. 
As a second example, we may require that we have selected a set of 
candidate inference models,
of which only the best should be used in the actual inference.
In the first example, we specified the condition $C_{i}$ that this
generative model should always be run, whereas in the second example,
we specified condition $C_{i}$ that a candidate model should only be run
when being the best.
A 'best' model is defined as the inference model with
the highest evidence (a.k.a. marginal likelihood), given the alignment 
simulated in the previous step.
The evidence of an inference model is estimated by nested 
sampling [\cite{maturana2018model}], using the \verb;NS; BEAST2 package. We note that scripted use of BEAST2 packages is only possible under Linux and Mac.
Windows systems can do the model comparison for shorter DNA sequences 
using the web interface of \verb;mcbette; [\cite{mcbette}].

The third step infers the posteriors,
using the simulated alignment (Fig.~\ref{fig:pipeline}, 2a $\rightarrow$ 4a),
and the inference models that were selected in the previous step (3). 
For each selected experiment a posterior is inferred, using the 
\verb;babette; [\cite{bilderbeek2018babette}] R package which makes use of BEAST2. 
This step usually takes up most of the full pipeline's computation time.

@@

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
We define a phylogeny $\tau$ is the combination of 
branching times $\Vec{t}$ and topology $\psi$.
Assuming that a non-standard generative prior $\mathit{p_{G}}$
produces a tree $\tau_{\mathit{G}}$,
with branching times $\Vec{t}_{\mathit{G}}$ and 
topology $\psi_{\mathit{G}}$, the twinning process T, in accordance with a chosen inference 
model $\mathit{I}$, creates a tree $\tau_{\mathit{I}}$
with branching times $\Vec{t}_{\mathit{I}}$ while preserving the 
topology $\psi_{\mathit{G}}$:

@@

The remaining differences in error distributions will be 
consequences of different phylogenies and tree priors for the two parallel 
pipelines: the main pipeline uses the tree $\tau_{\mathit{G}}$ and a
possibly mismatched prior $\mathit{G}$, where the twin pipeline
uses the tree $\tau_{\mathit{I}}$ for a matching prior $\mathit{I}$.
So, for the twin trees, if the process is repeated a sufficient number of times, the resulting error distribution cannot be 
attributed to a mismatch between the true tree prior and the tree prior used in
inference.

@@

\verb;pirouette; will be made available on CRAN from which it can then be easily installed easily:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install.packages("pirouette")
\end{lstlisting}

@@

\begin{lstlisting}[
    language = R,
    floatplacement = ht,
    frame = single
]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}
To start using \verb;pirouette;, load its functions in the global namespace first:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
library(pirouette)
\end{lstlisting}
Because \verb;pirouette; calls BEAST2, BEAST2 must be installed. 
This can be done from within R, using:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2()
\end{lstlisting}
For the option to select the best candidate model,
\verb;pirouette; needs the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:
\begin{lstlisting}[language=R, floatplacement=ht, frame=single]
install_beast2_pkg("NS")
\end{lstlisting}

An overview of \verb;pirouette;'s main functions is shown in 
Table~\ref{tab:functions}. 
Their usage is demonstrated in the example code below.
All \verb;pirouette;'s functions are documented,
have a useful example and sensible defaults.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 mononucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

As the site and clock models used here are also the defaults, 
the function arguments can be safely omitted: we just show these explicitly for clarity.

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    ERQ 1: the inference error made 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, and hence 
usually it is not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is what standard tree prior should be 
used in the inference.
There are two ways to deal with this problem: hand-picking a tree
prior, or picking the best of a set of tree priors. 
Hand-picking a tree prior can be rational if one standard tree prior 
is more suitable than others for the inference, for example, when
the novel tree prior is most similar to a specific standard tree prior.
Instead of hand-picking, the best inference model from a list of candidate 
models can also be used in inference. Selecting the 'best' model in
\verb;pirouette; means selecting the model with the highest evidence (also known as marginal likelihood).

@@

In the second step, we deviate from the answer to ERQ 1 by providing a 
set of experiments $\{\mathit{X_{i}}\}_{i=1}^{40}$. All the experiments have 
the same inference conditions $\mathit{C_{i}} = \mathit{C}$ but different 
candidate inference models $\mathit{I_{i}}$. Only one of them will be used 
to run the actual full inference. According to the specified $\mathit{C}$, 
\verb;pirouette; will measure the evidence for each of them and select only 
the one with the highest evidence as the best candidate.

As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function for this choice:
\verb;create_all_experiments; will create a full set of 40 experiments,
containing the inference models of all combinations of 
4 site models, 2 clock models and 5 tree priors.

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
as shown in Figure~\ref{fig:example_2}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    ERQ 2: the inference error 
    when selecting the best inference model.
  }
  \label{fig:example_2}
\end{figure}

@@

\begin{lstlisting}[
  language = R,
  floatplacement = ht,
  frame = single,
  label = {lst:create_twinning_params},
  caption = Create the default twinning parameters.
]
twinning_params <- create_twinning_params(
  twin_model = "bd", 
  method = "random_tree"
)
\end{lstlisting}
\giovanni{We want to import the "twin\_model" from the inference model.}
\richel{Yes, that is an option, but according to Rampal not important enough
to add now}
\rampal{did I say that? I think it's OK not to do it in the paper, but perhaps it can be noted.}

@@

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}),
The output is shown in Figure~\ref{fig:example_3}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    ERQ 3: the inference error made 
    when hand-picking a generative tree prior compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

@@

Rather than hand-picking a tree prior, 
one can select the best model from a set,
as in ERQ 2.
The resulting error distribution contains the full error accumulated in the pipeline, To quantify to what extent this error is due to the mismatch between 
the generative model $\mathit{G}$ and the selected inference model 
$\mathit{I}$, we can use twinning (see section \ref{subsec:twinning}: 'Twinning') to provide the baseline error.

@@

We specify that we want all 40 inference models to compete 
by calling \verb;create_all_experiments;,
as we previously did in listing~\ref{lst:create_all_experiments}.
We use again the default error measurement setup,
as shown in listing~\ref{lst:create_error_measure_params}.
To get an idea of a baseline error made by BEAST2, 
we also make use of the twinning option by using
the same call to \verb;create_pir_params; as shown in 
listing~\ref{lst:create_pir_params_with_twinning}.
We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_4}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    ERQ 4: the inference error made 
    when selecting the 'best' tree prior compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_4}
\end{figure}

@@

We use again the default error measurement parameters
created in listing~\ref{lst:create_error_measure_params}.
We enable twinning, by calling \verb;create_pir_params;
as shown in listing~\ref{lst:create_pir_params_with_twinning}.

We run \verb;pirouette; (listing~\ref{lst:pir_run}) 
and plot the results (listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_5}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the inference error made, 
    when the generative model is hand-picked, 
    compared to the best of 39 candidate models, 
    each in relationship to the background error measured from the twin tree.
  }
  \label{fig:example_5}
\end{figure}

@@

In the five example research questions, 
we showed how to use \verb;pirouette; 
to quantify the importance of a tree prior in Bayesian phylogenetics.
In all ERQs we used the most simple generative tree prior possible,
which is the Yule tree prior.
In principle any other (more complex) generative tree prior 
can be tested, but we chose to provide the 
simplest (and fastest to run) examples.
All tests were performed on a single original tree, 
hence not providing us enough statistical power to 
evaluate the level of accuracy of the inference.
